Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-24 submission/manuscript.Rmd
drafting fig 4 results

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.

>>>>

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.

### KDM5C-mediated removal of H3K4me3 during early embryogenesis promotes long-term repression of germline genes

**Figure 4: Loss of KDM5C's catalytic activity impairs DNAme placement and long-term silencing of germline genes**
	A) KDM5C binding in EpiLCs vs PNCs - number of peaks and percentage of bound genomic regions
	B) KDM5C binding in Epilcs vs pncs germline genes - GO 
	C) Left - bar graph of germline DEGs KDM5C is bound to in EpiLCs Right - Example bigwigs
		* Brain and EpiLC shared, EpiLC-specific, Brain-specific, 
	D) Bigwigs of Increase in H3K4me3 in Kdm5c-KO amygdala at germline genes
	E) Increase in H3K4me2 in EpiLCs at germline genes (left - bigwigs, right average signal at germline gene TSS? Just KDM5C-bound genes?)
	* Previous studies only looked at ESCs, unknown if catalytic activity is required for long-term repression, especially since DNA methylation is placed later). 
	F) KDM5C expression ESC --> EpiLC (increasing then decreasing) Left - RNA right - protein 
	G) RNA expression of germline genes with catalytic dead rescue (Ilakkiya)
	H) DNA methylation in WT and 5CKO EpiLCs (Ilakkiya)

Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription. However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including master regulators of meiosis and germline identity. While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other master regulators that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.

To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: 2,437, PNCs: 1,610, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A). Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B). Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).

We then assessed KDM5C binding at germline genes expressed _Kdm5c_-KO cells to assess if their dysreuglation is due to direct loss of KDM5C binding during embryogenesis.

	* KDM5C was bound to the promoter of many genes dysregulated in both EpiLCs and brain datasets, including _D1Pas1_ and _Cyct_.
	* KDM5C was not bound to many of the brain-specific germline DEGs, such as XXX.
	* While KDM5C binds to many germline DEGs in EpiLCs, including _Dazl_, many were also unbound. This included the germline transcription factor _Stra8_.
	* All genes lacked binding in PNCs.
	* Altogether, these data indicate that while some germline genes are misexpressed due to direct loss of KDM5C binding during emryogenesis, secondary downstream mechanisms can also promote their aberrant transcription.

----

Although KDM5C is generally thought to supress transcription through erasure of histone 3 lysine 4 di- and trimethylation (H3K4me2/3), recent studies in ESCs have suggested KDM5C's repression _Dazl_ is independent of its catalytic activity[@guptaGenomewideScreenReveals2023]. It is currently unclear if KDM5C's catalytic activity aids in the long-term repression of germline genes, such as enabling the placement of DNA methylation.

* In congruence with previous work in the _Kdm5c_-KO hippocampus[@scandagliaLossKdm5cCauses2017], we observed ectopic H3K4me3 peaks in the _Kdm5c_-KO amygdala at the transcription start site (TSS) of germline genes (Figure 4).
* We additionally observed a marked increase in H3K4me2 at germline gene promoters in _Kdm5c_-KO EpiLCs.
	* This increase was greatest at germline genes directly bound by KDM5C (?)
